Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-01-29 analysis/paper/paper.Rmd
started incorporating TBs awesome edits

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall, biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
keywords: |
  Northern Pikeminnow; Chinook Salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindering population recovery in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia rivers have created favorable conditions for fishes known to consume juvenile salmonids, including the Northern Pikeminnow *Ptychocheilus oregonsis*. Indeed, Northern Pikeminnow have been estimated to consume 8% of the approximately 200 million juvenile salmonids emigration through those rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above these reservoir systems. In this study, we examine predation by Northern Pikeminnow on juvenile salmonids, focusing on Chinook Salmon *Oncorhynchus tshawytscha*, in a slow-water reach of the Salmon River, Idaho known as Deadwater Slough. The reach contains historically important overwinter rearing and spring emigration habitat for Chinook Salmon and other ESA-listed species, and was formed by the failure of a small mining dam in a reservoir in the late 19th century. We estimate the abundance of Northern Pikeminnow in Deadwater Slough during peak emigrations of Chinook salmon, and further, use a bioenergetics approach to estimate the consumption potential of the Northern Pikeminnow population to assess potential impacts to Chinook Salmon emigrants. Finally, we evaulate potential impacts to adult returns of local populations. ***We found that... (insert concluding findings here.*** It is likely that predation of juvenile Chinook Salmon in Deadwater Slough has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG hindering local recovery efforts.

@@

The Snake and Columbia rivers drain a network of inland tributaries that provide essential spawning and rearing habitat for anadromous Pacific salmon *Oncorhynchus* spp. and steelhead *O. mykiss*. These fishes were historically abundant throughout the basin and consequently have tremendous ecological, cultural, and economic value. Over the last century, salmonid stocks throughout the Snake and Columbia rivers have undergone significant declines related to habitat modification in the freshwater environment (e.g., removal of beavers, river channel simplification, water withdrawals, hydrosystem development) and poor marine conditions, both affecting their survival and recruitment at multiple life stages. Consequently, action agencies have made considerable investments in the rehabilitation of tributary ecosystems and evaulation of factors attributed to the species decline [@Roni2018].

One potentially important, but perhaps under-appreciated source of salmonid mortality is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of population recovery [@Fresh2003; @Winther2020]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations, including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency. Because the Deadwater Slough supports rearing for DSR migrants and is part of the migration corridor for NRR migrants, and juveniles from all Upper Salmon populations must rear in or migrate through it, it represents a significant habitat for the Upper Salmon MPG.

Recent studies have examined the downstream movement, distribution, and (apparent) survival of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach compared to adjacent reaches [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had a 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during the fall and early winter months [@Ackerman2018; @Porter2019]. Low survival in these cases was attributed to low-velocity water delaying movement rates, thereby increasing predation risk.

In this study, we estimated the abundance of Northern Pikeminnow in the Deadwater Slough and evaluated potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. We hypothesize that high densities of piscivorous predators in the Deadwater Slough may explain the reduced survival (or apparent survival) observed for juvenile Chinook Salmon and Sockeye Salmon. To test this, our objectives were three-fold:

1. Estimate the abundance (or relative abundance) of potential predators in the Deadwater Slough during the peaks of fall (DSR) and spring (NRR) juvenile emigrations.
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage.
3. Use a bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population at Deadwater Slough to assess their potential impacts to DSR and NRR Chinook Salmon emigrants.

We also consider how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon MPG.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located approximately 6 river km downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973] forming an alluvial fan. The fan created a slow, deep section in the river, spanning approximately `r round(ddw_acres)` acres (`r round(ddw_hectares)` hectares), averaging `r round(ddw_avg_width_m)` m width, and resembling a small, narrow pond or reservoir.

## Northern Pikeminnow Population Size

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our initial intent was to sample during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span, but due to budget constraints the spring 2021 effort was limited to a single week. Fall efforts were constrained within a two-week window to minimize predator emigration/immigration and to constrain the population to meet assumptions of the mark-recapture models.

Multiple capture methods were employed during the initial fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were then abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while also minimizing potential impacts to ESA-listed adult steelhead present in the Salmon River during our sampling efforts.<!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted. During the fall 2019 effort it additionally became clear that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is our focus.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. During a sampling occassion (day), anglers would boat or hike their catch (periodically or upon filling a livewell) to a processing station at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and released. Fish that died prior to release were still counted as captured (or recaptured), but not included in the number of marked fish available for recapture after that date. Released individuals were distributed throughout Deadwater Slough to facilitate mixing back into the population. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to facilitate calculations of CPUE.

@@

Four different mark-recapture estimators were employed to explore variation in abundance estimates among them. Estimators fell into two broad categories: single census and multiple census. For the single census estimators, we treated the first week as the first (mark) sample, and the following (second) week as the second (recapture) sample, pooling data within each of those weeks. Alternatively, the multiple census estimators treat each day as a sample and use information about the total marked fish from all previous samples to infer the total abundance.

For the single census estimators, we first used the Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M)(n)}{(m)}
$$
where $M$ is the total number of fish marked and returned to the population during the first sample (week), $n$ is the total number of fish caught in the second sample (week), and $m$ is the number of marked fish caught during the second week. Because the Lincoln-Peterson estimator can be biased with small samples, we also employed the Chapman-modified Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
For a multiple census estimator, we first utilized the Schnabel method:

@@

where $M$, $n$, and $m$ are indexed by each sample (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999]. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Champman modification). Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and 2020 sample efforts (the estimate is of the initial population size). However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall efforts to the total CPUE in the spring efforts, then multiplied that ratio by the average abundance from the fall efforts. This approach assumes equal capture probabilities between fall and spring efforts.

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) sample(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken.

@@

Northern Pikeminnow in Deadwater Slough were not only abundant, but anglers reported that many were large-bodied. For that reason, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" (larger) individuals and "stock-sized" individuals:

@@

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. Quality and stock size definitions vary by species. We calculated the PSD for Northern Pikeminnow in Deadwater Slough using `r prettyNum(q_size, digits = 0)` mm TL for quality-size and `r prettyNum(s_size, digits = 0)` mm TL for stock-size [@Winther2020].

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as that portion consisting of fish parts or matter. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones, or were categorized as unknown [@Hansel1988; @Frost2000]. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

>>>>

---
title: "Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall, biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
keywords: |
  Northern Pikeminnow; Chinook Salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindering population recovery in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia rivers have created favorable conditions for fishes known to consume juvenile salmonids. The native Northern Pikeminnow are one of these fishes and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating throu the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above these reservoir systems. In this study, we examine predation by Northern Pikeminnow on Chinook Salmon in a slow-water reach of the Salmon River, Idaho known as Deadwater Slough. The reach was (at least partially) formed by the failure of a small mining dam in a nearby drainage during the late 19th century and contains historically important overwinter rearing habitat is a migration corridor for Chinook Salmon and other Endangered Species Act-listed species. We estimate the abundance of Northern Pikeminnow during peak emigrations to be ***insert numbers here***. Using a bioenergetics approach we estimate the consumption potential of the Northern Pikeminnow population to be ***key results here***. This resulted in the potential consumption of ***numbers*** Chinook Salmon emigrants. Using a sensitivity analysis, we evaluated the potential impact to adult returns of local Chinook Salmon populations to be between ***number with CIs***. It is likely that predation of juvenile Chinook Salmon in Deadwater Slough has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River major population group, thereby hindering local recovery efforts.

@@

The Snake and Columbia rivers drain a network of inland tributaries that provide essential spawning and rearing habitat for anadromous Pacific salmon *Oncorhynchus* spp. and steelhead *O. mykiss*. These fishes were historically abundant throughout the basin and consequently have tremendous ecological, cultural, and economic value [@Atlas2021; @Lewis2019]. Over the last century, salmonid stocks throughout the Snake and Columbia rivers have undergone significant declines related to habitat modification in the freshwater environment (e.g., removal of beavers, river channel simplification, water withdrawals, hydrosystem development) and poor marine conditions, affecting their survival and recruitment at multiple life stages [@Crozier2020; @Clark2020; @Justice2017]. Consequently, action agencies have made considerable investments in the rehabilitation of tributary ecosystems and evaluation of factors attributed to the species decline [@White2021; @Roni2018].

One potentially important, but perhaps under-appreciated source of salmonid mortality is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of population recovery [@Fresh2003; @Winther2020]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume ~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is in a river reach containing historically important overwinter rearing habitat, includes the emigration corridor for juvenile Chinook Salmon *O. tshawytscha*, and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations, including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is an intensively monitored watershed to determine responses to restoration actions [@Zimmerman2012]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for the Upper Salmon MPG as is supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

Recent studies have examined the downstream movement, distribution, and apparent (hereafter 'survival') of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach compared to adjacent reaches [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had a 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during the fall and early winter months [@Ackerman2018; @Porter2019]. Low survival was attributed to low-velocity water delaying movement rates, thereby increasing predation risk.

In this study, we estimated the abundance of Northern Pikeminnow in the Deadwater Slough and evaluated potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. We hypothesize that high densities of piscivorous predators in the Deadwater Slough may explain the reduced survival observed for juvenile Chinook Salmon. To test this, our objectives were:

1. Estimate the abundance of potential predators in the Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3. Use a bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4. Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow impacts adult returns to the Upper Salmon MPG.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (RKM) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event and upslope instability that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973]. That sediment formed an alluvial fan which created a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and resembling a small reservoir.

## Northern Pikeminnow Demographics

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Northern Pikeminnow were surveyed near the peaks of the fall DSR and spring NRR emigrations. The initial intent was to survey during the emigration of the 2018 brood year (fall 2019 and spring 2020), however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until fall 2020. Although this allowed an additional survey during the fall DSR emigration, surveys during the spring NRR emigration were still lacking. Consequently, an additional week of surveying was added in spring 2021. Fall surveys occurred over two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present in the Salmon River during our surveys.<!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. Each survey day, anglers boated or hiked their catch (periodically or upon filling a livewell) to a processing station at the boat ramp, approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and then were released. Released fish were distributed throughout Deadwater Slough to facilitate mixing back into the population. Fish that died prior to release were included in the total capture (or recapture) count, but not included in the number of marked fish available for recapture after that date. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to calculate CPUE as number of Northern Pikeminnow caught per angler hour.

@@

Two types of mark-recapture estimators were used to explore variation in abundance estimates: a single census and a multiple census estimator. Estimators fell into two broad categories: single census and multiple census. For the single census estimator, we treated the first week of the survey as the mark event and the following week as the recapture event, pooling data within each event week. Alternatively, the multiple census estimator treated each day as an event and use information about the total number of marked fish from all previous events to calculate the total abundance.

For the single census estimator, we used the Chapman-modified Lincoln-Peterson estimator:

<!--
$$
\hat{N} = \frac{(M)(n)}{(m)}
$$
-->
$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$

where $M$ is the total number of fish marked and returned to the population during the first event (week), $n$ is the total number of fish caught in the second event (week), and $m$ is the number of marked fish caught during the second week. For a multiple census estimator, we used the Schnabel method:

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999]. The Schnabel estimator is essentially a weighted average of a series of Chapman-modified Lincoln-Peterson estimators.

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those locales, marked fish may have been less susceptible to capture. To account for this potential bias, we adjusted the Schnabel estimator with an assumption that marked and released fish during each survey day were not available for recapture for 48 hours.

<!-- delayed mixing Schnabel equation here; include statement about fish being "hook shy" -->

<!--
Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$
-->

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and 2020 surveys. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall survey to the total CPUE in the spring survey, then multiplied that ratio by the average abundance from the fall surveys.

$$
\hat{N}_{spring} = \overline{N}_{fall} * \frac{CPUE_{spring}}{CPUE_{fall}}
$$

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into a linear density by dividing by the length of Deadwater Slough (km) measured in GIS software and using drone-generated orthography.

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken.

@@

Anglers reported that Northern Pikeminnow in Deadwater Slough were large-bodied. To describe the population size structure, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough. PSD is a ratio, expressed as a percentage, between the number of "quality-sized" (larger) individuals and "stock-sized" individuals:

@@

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. For Northern Pikeminnow in Deadwater Slough, we used `r prettyNum(q_size, digits = 0)` mm TL for quality-length and `r prettyNum(s_size, digits = 0)` mm TL for stock-length [@Winther2020].

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks  to be analyzed in a controlled environment. For each stomach sample, wet weight (grams) was recorded for the total stomach contents, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones, or were categorized as unknown [@Hansel1988; @Frost2000]. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology.
